---
title: "2019 Bootcamp Report 1"
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---
Batch effect can be a cause of lack of reproducibility. Here we demonstrate batch effect with the data set in the study from Spielman group. [^1]

[^1]:Spielman, R. et al. Common genetic variants account for differences in gene expression among ethnic groups. *Nature Genetics*. **39**, pages 226–231 (2007)

```{r message=FALSE, warning=FALSE}

# load libraries and datasets
# source("http://bioconductor.org/biocLite.R")
# biocLite("Biobase")
# biocLite("genefilter")
# devtools::install_github("genomicsclass/GSE5859")

library(Biobase)
library(rafalib)
library(genefilter)
library(GSE5859)
data(GSE5859)
```
```{r }
# gene expression levels  
geneExpression = exprs(e)
```

# Data collection dates
```{r }
# print date info
sampleInfo = pData(e)
head(sampleInfo$date)
```

Check years that the data from different ethnic groups were collected. We can immediately see that year and ethnicity are almost completely confounded.
```{r }
# Output: print a table of year vs ethnic groups
year = factor( format(sampleInfo$date,"%y") )
tab = table(year,sampleInfo$ethnicity)
print(tab)
```
# Differential expression analysis
```{r }
##remove control genes
out <- grep("AFFX",rownames(geneExpression))

```
- We first use a t-test to perform differential expression analysis between CEU and ASN.

```{r }
eth <- sampleInfo$ethnicity
ind<- which(eth%in%c("CEU","ASN"))
res1 <- rowttests(geneExpression[-out,ind],droplevels(eth[ind]))

```
- We then perform the same analysis between 2002 vs 2003 within CEU.

```{r }
ind <- which(year%in%c("02","03") & eth=="CEU")
res2 <- rowttests(geneExpression[-out,ind],droplevels(year[ind]))

```
- We plot histograms for p-values as in Akey paper. 
```{r }
mypar(1,2)
hist(res1$p, main="p-value of CEU vs ASN")
hist(res2$p, main="p-value of 2002 vs 2003 for CEU")
```

